---
title: "NBA longitudinal efficiency analysis"
author: "Will Tackett, Jacob Vidergar"
date: "2024-12-15"
output: pdf_document
---

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(performance)
library(RColorBrewer)
library(sjPlot)
library(table1)
```

## Exploratory Data Analysis
```{r}
data <- read.csv("all_seasons.csv")
data <- data %>% select(-X)
head(data)
```

```{r}
summary(data)
str(data)
```

### Filter data
```{r, warning=FALSE}
filtered_data <- data %>%
  # Add a column for games played in the season
  mutate(gp_season = case_when(
    season == "1998-99" ~ 50,
    season == "2011-12" ~ 66,
    TRUE ~ 82
  )) %>%
  # adjust naming of draft round categories
  mutate(draft_round_combined = case_when(
    draft_round == 0 ~ "Undrafted", 
    draft_round == "Undrafted" ~ "Undrafted", 
    draft_round %in% c(1, 2) ~ as.character(draft_round),
    TRUE ~ NA_character_
  )) %>%
  # Filter for relevant draft round categories
  filter(draft_round_combined %in% c("Undrafted", "1", "2")) %>%
  # Calculate the percentage of games played
  mutate(gp_pct = gp / gp_season) %>%
  # Filter for players who played at least 50% of games
  filter(gp_pct > 0.5)

# make season continuous
filtered_data <- filtered_data %>%
  mutate(season_continuous = as.numeric(substr(season, 1, 4)))

# Create a new column combining 2 and Undrafted into one category
filtered_data <- filtered_data %>%
  mutate(draft_round_combined_new = ifelse(draft_round_combined %in% c("2", "Undrafted"), 
                                           "2_or_Undrafted", 
                                           draft_round_combined))
```

```{r}
# convert important variables from chr to factor
filtered_data$draft_round_combined <- as.factor(filtered_data$draft_round_combined)
# filtered_data$season <- as.factor(filtered_data$season)
filtered_data$player_name <- as.factor(filtered_data$player_name)
```

```{r}
# set min threshold for games played (50%)
print("before threshold")
# Count the number of observations for each level
table(filtered_data$draft_round_combined)

print("after threshold")
# Keep observations where gp / gp_season >= 0.5
filtered_data <- filtered_data %>%
  filter(gp / gp_season >= 0.5)

# Count the number of observations for each level
table(filtered_data$draft_round_combined)
```

```{r}
# Prepare the data for Table 1
table_data <- filtered_data %>%
  select(draft_round_combined, ts_pct, player_height, player_weight, age, gp_pct) %>%
  mutate(draft_round_combined = factor(
    draft_round_combined,
    levels = c("1", "2", "Undrafted"),
    labels = c("1st Round", "2nd Round", "Undrafted")
  ))

# Apply variable labels using label()
label(table_data$draft_round_combined) <- "Draft Round"
label(table_data$ts_pct) <- "True Shooting Percentage"
label(table_data$player_height) <- "Height (inches)"
label(table_data$player_weight) <- "Weight (lbs)"
label(table_data$age) <- "Age"
label(table_data$gp_pct) <- "Games Played (%)"

# Create the Table 1
table1(
  ~ ts_pct + player_height + player_weight + age + gp_pct | draft_round_combined,
  data = table_data,
  overall = "Overall",
  render = function(x, name, ...) {
    if (is.numeric(x)) {
      # Customize numeric summaries: Mean (SD)
      sprintf("%0.2f (%0.2f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))
    } else {
      # Show counts for categorical variables
      table1::render.default(x, name, ...)
    }
  }
)

``` 

### Preliminary Plots
```{r}
# Average `ts_pct` by the new combined draft_round category and season
draft_round_ts_pct_combined <- filtered_data %>%
    group_by(draft_round_combined, season) %>%
    summarize(avg_ts_pct = mean(ts_pct, na.rm = TRUE))

# Ensure `season` is treated as a factor with the correct order
draft_round_ts_pct_combined$season <- factor(draft_round_ts_pct_combined$season, levels = sort(unique(draft_round_ts_pct_combined$season)))
```


```{r}
ggplot(filtered_data, aes(x = season, y = ts_pct, color = draft_round_combined_new, group = draft_round_combined_new)) +
  stat_summary(fun = mean, geom = "line", size = 1) + # Line plot for mean TS% by season
  stat_summary(fun = mean, geom = "point", size = 2) + # Points for clarity
  labs(
    title = "Interaction Effect of Draft Round and Season on TS%",
    x = "Season",
    y = "True Shooting Percentage",
    color = "Draft Round"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels
```

```{r}
ggplot(filtered_data, aes(x = draft_round_combined_new, y = ts_pct, fill = draft_round_combined_new)) +
  geom_boxplot() +
  labs(
    title = "Distribution of True Shooting Percentage by Draft Round",
    x = "Draft Round",
    y = "True Shooting Percentage"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
ggplot(filtered_data, aes(x = player_height, y = player_weight, color = ts_pct)) +
  geom_point(alpha = 1, size = 3) +
  scale_color_viridis_c(
    name = "TS%", 
    option = "plasma",        # Use a vibrant "plasma" color scheme
    limits = c(0.45, 0.7),    # Adjust the range for TS%
    breaks = seq(0.45, 0.7, by = 0.05)  # Add meaningful breaks
  ) +
  labs(
    title = "Height, Weight, and True Shooting Percentage",
    x = "Player Height (inches)",
    y = "Player Weight (lbs)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, hjust = 0.5),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```


## Modeling
```{r}
# linear_model <- lm(ts_pct ~ draft_round_combined, data = filtered_data)
# summary(linear_model)

# Fit a mixed effects model where:
# - `season` is modeled as a fixed effect to assess the change in `ts_pct` over time.
# - `draft_round_combined_new` is modeled as a fixed effect to adjust for round differences.
# - A random intercept for `player_id` to account for player-specific variability.

model <- lmer(ts_pct ~ season * draft_round_combined + player_height + player_weight + age + (1 | player_name), 
              data = filtered_data)

# View the model summary
summary(model)
```

Try season as continuous variable
```{r}
filtered_data <- filtered_data %>%
  mutate(season_continuous = as.numeric(substr(season, 1, 4)))

model <- lmer(ts_pct ~ season_continuous * draft_round_combined + player_height + player_weight + age + (1 | player_name), data = filtered_data)

# View the model summary
summary(model)
```

Histogram for age
```{r}
# Basic histogram
hist(filtered_data$age,
     breaks = 10, # Number of bins
     col = "skyblue", # Bar color
     border = "black", # Border color
     main = "Age Distribution", # Title
     xlab = "Age", # X-axis label
     ylab = "Frequency", # Y-axis label
     ylim = c(0, 2000)) # Adjust Y-axis limits for better view

# Add a grid for better readability (optional)
grid()

```

Recode age based on histogram
```{r}
# Create a new categorical variable based on age
filtered_data$career_stage <- cut(filtered_data$age,
                                  breaks = c(-Inf, 25, 30, Inf), # Define age ranges
                                  labels = c("Rookie", "Mid-Career", "Veteran"), # Labels
                                  right = TRUE) # Include the upper boundary in each interval

# Check the distribution of the new variable
table(filtered_data$career_stage)

```

New model with career_stage instead of age
```{r}
model <- lmer(ts_pct ~ season_continuous * draft_round_combined + player_height + player_weight + career_stage + (1 | player_name), data = filtered_data)

# View the model summary
summary(model)
```

Residual Diagnostics
```{r}
# Extract residuals
residuals <- residuals(model)

# Plot residuals vs. fitted values
plot(fitted(model), residuals,
     main = "Residuals vs. Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0, col = "red", lwd = 2)

# Check normality of residuals
qqnorm(residuals, main = "Q-Q Plot of Residuals")
qqline(residuals, col = "red", lwd = 2)

# Histogram of residuals
hist(residuals, breaks = 30, main = "Histogram of Residuals",
     xlab = "Residuals", col = "lightblue", border = "black")

```

Random Effects Diagnostics
```{r}
# Extract random effects
ranef_data <- as.data.frame(ranef(model)$player_name)

# Plot random effects
ggplot(ranef_data, aes(x = `(Intercept)`)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  labs(title = "Random Effects (Player Intercept)", x = "Random Intercept", y = "Frequency") +
  theme_minimal()

```

Multicollinearity Check
```{r}
# Using performance package
# Check collinearity for your mixed-effects model
check_collinearity(model)
```

Potential Logit Transform
```{r}
filtered_data$ts_pct_logit <- log(filtered_data$ts_pct / (1 - filtered_data$ts_pct))
model_logit <- lmer(ts_pct_logit ~ season_continuous * draft_round_combined +
                      player_height + player_weight + career_stage + (1 | player_name), data = filtered_data)

summary(model_logit)
```


Compare models
```{r}
AIC(model, model_logit)
```

Non linear effects of height and weight?
```{r}
# Quadtratic terms
model_poly <- lmer(ts_pct ~ season_continuous * draft_round_combined + 
                     poly(player_height, 2) + poly(player_weight, 2) + 
                     career_stage + (1 | player_name), data = filtered_data)
summary(model_poly)

# Spline model
library(splines)
model_spline <- lmer(ts_pct ~ season_continuous * draft_round_combined + 
                       ns(player_height, df = 3) + ns(player_weight, df = 3) + 
                       career_stage + (1 | player_name), data = filtered_data)
summary(model_spline)

```

Compare models
```{r}
anova(model, model_poly, model_spline)
AIC(model, model_poly, model_spline)
```

```{r}
library(sjPlot)
plot_model(model_poly, type = "pred", terms = "player_height [all]")
plot_model(model_poly, type = "pred", terms = "player_weight [all]")

# Group by player_height and calculate the mean ts_pct
avg_ts_by_height <- filtered_data %>%
  group_by(player_height) %>%
  summarize(avg_ts_pct = mean(ts_pct, na.rm = TRUE))

# View the summarized data
head(avg_ts_by_height)

ggplot(avg_ts_by_height, aes(x = player_height, y = avg_ts_pct)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "darkblue") +
  labs(
    title = "Average TS% by Player Height",
    x = "Player Height",
    y = "Average TS%"
  ) +
  theme_minimal()

# Group by player_weight and calculate the mean ts_pct
avg_ts_by_weight <- filtered_data %>%
  group_by(player_weight) %>%
  summarize(avg_ts_pct = mean(ts_pct, na.rm = TRUE))

# View the summarized data
head(avg_ts_by_weight)

ggplot(avg_ts_by_weight, aes(x = player_weight, y = avg_ts_pct)) +
  geom_line(color = "red", size = 1) +
  geom_point(color = "darkred") +
  labs(
    title = "Average TS% by Player Weight",
    x = "Player Weight",
    y = "Average TS%"
  ) +
  theme_minimal()

```



LeBron James sub-analysis
```{r}
# Filter the data for LeBron James
leData <- subset(filtered_data, player_name == "LeBron James")

# Plot ts_pct over season_continuous for LeBron James
plot(leData$season_continuous, leData$ts_pct,
     type = "o", # Line plot with points
     col = "blue", # Line color
     pch = 16, # Point style
     lwd = 2, # Line width
     xlab = "Season (Continuous)", # X-axis label
     ylab = "True Shooting Percentage (ts_pct)", # Y-axis label
     main = "LeBron James: TS% Over Time") # Title

# Add grid lines for better readability (optional)
grid()
# Calculate the league average ts_pct for each season
league_avg <- aggregate(ts_pct ~ season_continuous, data = filtered_data, mean)

# Plot league average ts_pct over season_continuous
plot(league_avg$season_continuous, league_avg$ts_pct,
     type = "o", # Line plot with points
     col = "red", # Line color
     pch = 16, # Point style
     lwd = 2, # Line width
     xlab = "Season (Continuous)", # X-axis label
     ylab = "League Average TS%", # Y-axis label
     main = "League Average TS% Over Time") # Title

# Add grid lines for better readability (optional)
grid()


leModel <- lm(ts_pct ~ career_stage, data = leData)

# View the model summary
summary(leModel)
```

